\documentclass[11pt,a4paper]{article}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{geometry}
\usepackage{hyperref}
\usepackage{listings}
\usepackage{color}
\usepackage{longtable}
\usepackage{booktabs}

\geometry{left=2cm,right=2cm,top=2cm,bottom=2cm}

\title{CS4031 - Compiler Construction\\Assignment 3: Bottom-Up Parser Design \& Implementation}
\author{Hussain Waseem Syed (22i-0893) \and Mohammad Abbas (22i-1409)\\Section C}
\date{\today}

\begin{document}

\maketitle
\tableofcontents
\newpage

\section{Introduction}
Bottom-up parsing is a technique used in compiler design to construct a parse tree for an input string beginning at the leaves (the bottom) and working up towards the root (the top). Unlike top-down parsing, which attempts to find a leftmost derivation, bottom-up parsing essentially traces a rightmost derivation in reverse via a shift-reduce mechanism. 

LR parsers (Left-to-right, Rightmost derivation) are the most powerful class of bottom-up parsers, capable of handling a very large class of context-free grammars. In this assignment, we implemented two variants: SLR(1) (Simple LR) and LR(1). SLR(1) builds its state machine without lookaheads (using LR(0) items) and uses global \texttt{FOLLOW} sets to decide when to perform reductions. This makes it simple but susceptible to conflicts in complex grammars. LR(1) embeds specific lookahead symbols into the parser states (LR(1) items), providing precise context for reductions, thereby resolving conflicts that SLR(1) cannot handle, at the cost of a larger number of generated states.

\section{Approach}
\subsection{Data Structures}
\begin{itemize}
    \item \textbf{Items}: 
    \begin{itemize}
        \item \texttt{LR0Item}: Represents an SLR item. Contains a reference to a \texttt{Production} and an integer \texttt{dotIndex} tracking the parsing progress.
        \item \texttt{LR1Item}: Composes an \texttt{LR0Item} and a \texttt{Set<String>} of specific lookahead symbols. 
    \end{itemize}
    \item \textbf{ItemSets (States)}: A state in the DFA, represented by a \texttt{Set} of items (either LR0 or LR1). Assigned a unique integer \texttt{stateId}.
    \item \textbf{ParsingTable}: A 2D structure mapping \texttt{(State, Symbol)} to an \texttt{Action}. Actions can be \texttt{SHIFT(state)}, \texttt{REDUCE(production)}, \texttt{ACCEPT}, or \texttt{ERROR}. Conflicts are tracked if an entry already exists when a new action is proposed.
    \item \textbf{Stack}: A unified parsing stack storing \texttt{(Symbol, State)} pairs, facilitating the shift-reduce trace.
\end{itemize}

\subsection{Algorithm Implementations (CLOSURE and GOTO)}
\textbf{CLOSURE}: For a given set of items $I$, we iteratively add items for any non-terminal immediately following the dot. For SLR(1), we just add \texttt{B $\rightarrow$ . $\gamma$}. For LR(1), we compute the new lookahead by taking the FIRST set of the symbols following $B$ combined with the current item's lookahead. This is implemented via a worklist algorithm that loops until no new items are added to $I$.

\textbf{GOTO}: For an item set $I$ and symbol $X$, we collect all items in $I$ where $X$ immediately follows the dot, advance the dot, and then compute the CLOSURE of this new set. The canonical collections are built by tracking a list of visited sets and systematically expanding unvisited sets via GOTO on all grammar symbols.

\subsection{Design Decisions and Trade-offs}
\begin{itemize}
    \item \textbf{Language Choice}: Java was selected for its robust object-oriented features, built-in Collection framework (essential for Set operations), and garbage collection.
    \item \textbf{Unified Table vs Split ACTION/GOTO}: We conceptually combined them into a single \texttt{ParsingTable} object, using maps for easy lookup, which also simplifies the formatted printing of the table.
    \item \textbf{Immutability}: Items are treated as immutable where possible to allow safe hashing and set comparisons during DFA construction.
\end{itemize}

\subsection{Lookahead Handling in LR(1)}
In LR(1), an item is of the form \texttt{[A $\rightarrow$ $\alpha$ . $\beta$, a]}. The lookahead `a' is propagated during the CLOSURE operation. If we have \texttt{[A $\rightarrow$ $\alpha$ . B $\beta$, a]}, we add \texttt{[B $\rightarrow$ . $\gamma$, b]} for each \texttt{b $\in$ FIRST($\beta$a)}. This logic is strictly enforced in \texttt{Items.java}. When building the parsing table, reductions are \textit{only} registered for the exact lookaheads stored in the \texttt{LR1Item}, as opposed to SLR(1) which indiscriminately reduces for all symbols in \texttt{FOLLOW(A)}.

\section{Challenges and Solutions}
\begin{itemize}
    \item \textbf{Challenge - Computing FIRST sets}: Dealing with $\epsilon$-productions and chains of nullable non-terminals.
    \item \textbf{Solution}: We implemented an iterative fixed-point algorithm that continuously passes through all productions, propagating FIRST sets and $\epsilon$-flags until no sets change size.
    \item \textbf{Challenge - Distinguishing shared items in LR(1) sets}: Creating a new item for every lookahead wastes memory and complicates equality checks.
    \item \textbf{Solution}: \texttt{LR1Item} maintains a \texttt{Set} of lookaheads. When merging sets during CLOSURE, if an item with the same core exists, its lookaheads are simply unioned together.
    \item \textbf{Challenge - Infinite loops during CLOSURE}: Left-recursive rules caused infinite loops during state generation.
    \item \textbf{Solution}: Implemented equality (\texttt{equals}) and hash code (\texttt{hashCode}) methods accurately for Items, leveraging \texttt{HashSet.contains()} to avoid reprocessing items that are already inside the closure set.
\end{itemize}

\section{Test Cases}
We tested 4 distinct grammars, with well over 5 input strings per grammar (combining valid and invalid cases).

\subsection{Grammar 1 (Simple Expressions)}
\textbf{Rules}: \texttt{Expr $\rightarrow$ Expr + Term | Term}, \texttt{Term $\rightarrow$ Factor}, \texttt{Factor $\rightarrow$ id}
\begin{itemize}
    \item \textbf{Valid Strings Tested}: 1. \texttt{id}, 2. \texttt{id + id}, 3. \texttt{id + id + id}, 4. \texttt{id + id + id + id}, 5. \texttt{id + id + id + id + id} (All accepted perfectly).
    \item \textbf{Invalid Strings Tested}: 1. \texttt{id * id}, 2. \texttt{( id )}, 3. \texttt{id +}, 4. \texttt{id + + id}, 5. \texttt{+ id}, 6. \texttt{+} (All rejected with precise syntax errors).
\end{itemize}

\subsection{Grammar 2 (Full Expressions with * and ())}
\textbf{Rules}: \texttt{Expr $\rightarrow$ Expr + Term | Term}, \texttt{Term $\rightarrow$ Term * Factor | Factor}, \texttt{Factor $\rightarrow$ ( Expr ) | id}
\begin{itemize}
    \item \textbf{Valid Strings Tested}: 1. \texttt{id}, 2. \texttt{id + id}, 3. \texttt{id * id}, 4. \texttt{id + id * id}, 5. \texttt{( id )}, 6. \texttt{( id + id ) * id}, 7. \texttt{( ( id + id ) * id + id )}, 8. \texttt{id + id + id + id}.
    \item \textbf{Invalid Strings Tested}: 1. \texttt{id +}, 2. \texttt{id + + id}, 3. \texttt{( id + id}, 4. \texttt{id * * id}, 5. \texttt{+ id}, 6. \texttt{( + )}.
\end{itemize}

\subsection{Grammar 3 (Classic SLR Conflict resolved by LR1)}
\textbf{Rules}: \texttt{Start $\rightarrow$ Lhs = Rhs | Rhs}, \texttt{Lhs $\rightarrow$ * Rhs | id}, \texttt{Rhs $\rightarrow$ Lhs}
\begin{itemize}
    \item \textbf{Valid Strings Tested}: 1. \texttt{id}, 2. \texttt{* id}, 3. \texttt{id = id}, 4. \texttt{id = * id}, 5. \texttt{* id = id}.
    \item \textbf{Invalid Strings Tested}: 1. \texttt{id =}, 2. \texttt{id = = id}, 3. \texttt{*}, 4. \texttt{id = id = id}, 5. \texttt{= id}.
    \item \textbf{Conflict Resolution}: SLR(1) generates a Shift/Reduce conflict (\texttt{s8/r5}) on \texttt{=} in State 4 because \texttt{=} is in \texttt{FOLLOW(Rhs)}. LR(1) resolves this because the lookahead restricts the reduction precisely. Thus strings 3, 4, and 5 are \textbf{rejected} by SLR(1) but \textbf{accepted} by LR(1).
\end{itemize}

\section{Comparison Analysis}
Below is the benchmarking data comparing SLR(1) and LR(1) for Grammar 3 (the conflict resolution grammar).

\begin{table}[h]
\centering
\begin{tabular}{|l|c|c|}
\hline
\textbf{Metric} & \textbf{SLR(1)} & \textbf{LR(1)} \\ \hline
Number of states & 10 & 14 \\ \hline
ACTION entries & 17 & 22 \\ \hline
GOTO entries & 7 & 9 \\ \hline
Total table entries & 24 & 31 \\ \hline
Build time (ms) & $\sim$4 ms & $\sim$2 ms \\ \hline
Has conflicts? & \textbf{YES} (Shift/Reduce) & \textbf{No} \\ \hline
Parsing Power & Weak (Global FOLLOW) & Strong (Item Lookahead) \\ \hline
Memory Usage & Lower & Higher \\ \hline
\end{tabular}
\caption{SLR(1) vs LR(1) Performance Profile (Grammar 3)}
\end{table}

\textbf{Analysis}: LR(1) is strictly more powerful. In Grammar 3, SLR(1) sees the item \texttt{Rhs $\rightarrow$ Lhs .} in state 4. It looks at \texttt{FOLLOW(Rhs)}, which contains \texttt{=}, so it decides to reduce on \texttt{=}. However, an assignment cannot start with a right-hand side. LR(1) resolves this because the exact lookahead for that specific item is only \texttt{\$}, preventing the erroneous reduction. The trade-off is memory usage: LR(1) generates 14 states compared to SLR(1)'s 10 states.

\section{Sample Outputs}

\subsection{Screenshots of Program Execution}
\textit{Note: As this is a textual report, the following listings represent the console execution output directly generated by the program.}

\begin{verbatim}
==============================================
 Running Grammar 3: Classic SLR(1) Conflict Example
==============================================
[String 3] tokens: [id, =, id]
  --- SLR(1) ---
Step  | Stack (symbol:state)       | Remaining Input   | Action
--------------------------------------------------------------------------------
1     | [$:0]                      | id = id $         | shift 2
2     | [$:0] [id:2]               | = id $            | reduce Lhs -> id
3     | [$:0] [Lhs:4]              | = id $            | CONFLICT: s8/r5
✘ String REJECTED
  Error: Syntax error at step 3: unexpected '=' in state 4
\end{verbatim}

\subsection{Item Sets and Parsing Tables}
\textbf{Sample LR(1) Item Sets (State 4 and 8 from Grammar 3):}
\begin{verbatim}
I4:
  Start -> Lhs • = Rhs, [$]
  Rhs -> Lhs •, [$]

I8:
  Start -> Lhs = • Rhs, [$]
  Rhs -> • Lhs, [$]
  Lhs -> • * Rhs, [$]
  Lhs -> • id, [$]
\end{verbatim}

\textbf{Sample LR(1) Parsing Table Excerpt:}
\begin{verbatim}
=== LR(1) Parsing Table ===
             ACTION       |      GOTO      
State | =  | *  | id | $  |Start|Lhs |Rhs |
-------------------------------------------------
I0    |    | s1 | s2 |    |  3  | 4  | 5  |
...
I4    | s8 |    |    | r5 |     |    |    |
I8    |    | s9 |s10 |    |     | 11 | 12 |
...
\end{verbatim}

\subsection{Parse Trees for Accepted Strings}
Parse tree successfully generated for \texttt{( id + id ) * id} in Grammar 2:
\begin{verbatim}
Expr
└── Term
    ├── Term
    │   └── Factor
    │       ├── (
    │       ├── Expr
    │       │   ├── Expr
    │       │   │   └── Term
    │       │   │       └── Factor
    │       │   │           └── id
    │       │   ├── +
    │       │   └── Term
    │       │       └── Factor
    │       │           └── id
    │       └── )
    ├── *
    └── Factor
        └── id
\end{verbatim}

\subsection{Conflict Examples and Resolutions}
In Grammar 3, SLR(1) hits a conflict in state 4 on input \texttt{=}. The ACTION table for SLR(1) in State 4 for column \texttt{=} is \texttt{s8/r5} (Shift 8 or Reduce 5). The program reports:
\texttt{CONFLICT DETECTED (1): Shift/Reduce conflict in state 4 on '=': existing='s8', new='r5'}
In LR(1), this is resolved. The table cleanly contains \texttt{s8} for State 4 on \texttt{=}. The reduction \texttt{r5} only happens on \texttt{\$} (EOF), as per the item lookahead.

\section{Conclusion: Lessons Learned and Insights}
This assignment highlighted the practical differences in power and complexity between SLR(1) and LR(1) parsing methodologies. We learned that while SLR(1) is conceptually simpler and yields smaller parsing tables, its reliance on global \texttt{FOLLOW} sets makes it too blunt for grammars with overlapping right contexts (like the L-value/R-value assignment grammar). 

LR(1) eliminates these spurious conflicts by tightly coupling contextual lookaheads to individual parser items. This precision significantly expands the class of grammars that can be parsed, albeit at the expense of computational and memory overhead to construct the larger state machine (e.g., table size expanded by ~30\% for our test cases). Implementing both from scratch solidified our understanding of DFA state generation, shift-reduce architecture, and automated AST (Abstract Syntax Tree) generation.

\end{document}
